On the effects of partying on next-day sports performance

### 2017-07-01 writing/draft.tex
Tidied introduction page.

\section{Introduction}
It's not unreasonable to think that in-game sporting performance can be affected 
partly by what takes place off the court. 
We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife; we call this a ``hangover effect''.
Part of the question is determining a reasonable way to measure
levels of nightlife, and correspondingly which cities are notorious for it; we colloquially
refer to such cities as ``party cities''.

\paragraph{Why do we care} 
Inquiring about this causal relationship is worthwhile for several reasons. The
NBA Commissioner Adam Silver has reported that there is \$400 billion per year spent 
on sports betting in the U.S. alone.\citep{asilver,cnnsilver} Apart from arbitrage opportunities,
knowing how player performance is affected by external factors could inform coaches of a more
optimal play strategy. E.g. if visitation to a city with active nightlife is correlated with
next-day injury, perhaps coaches might bench their star players if the latter game isn't crucial.

\paragraph{What do we do}
We exploit data on bookmaker spreads, the expected score 
differential between two teams after conditioning on observable performance. 
We expect a team to meet the spread half the time, since this is one of the easiest ways for 
bookmakers to guarantee a profit.\citep{dubnerlevitt} We construct a model which attempts to 
estimate the causal effect of  visiting a ``party-city'' on subsequent day 
performance as measured by the spread. In particular, we are only looking at
cases where games are played back-to-back within 24 hours of each other.
To the extent that next-day opponent is uncorrelated with exposure to treatment, 
we have identification in our variable of interest.

\paragraph{What do we find}
We replicate this analysis for both the National Basketball Association
(NBA) and Major League Baseball (MLB). In particular, we find that visiting
a city with active nightlife the day prior to a game does have a statistically
significant negative effect on a team's likelihood of meeting the point-spread set
by a betting house, for both NBA and MLB. Within the NBA, we analyze specific performance
metrics such as number of points allowed and number of rebounds obtained, and find that both are negatively effected by previous day visitation to a city with active nightlife. Having established a causal model, we realize a profitable betting scheme for MLB. We also perform a robustness check and see that if players rest more than
 24 hours after visiting a ``party-city'', the hangover effect dissipates.

@@

\subsection{Anecdotal evidence of the effects on sporting performance}
Professional players openly admit they are interested in nightlife: Matt Barnes recently complained that 
``there's no nightlife in Utah'' when asked about playing Jazz in the second round of NBA playoffs 
in 2017.\citep{cestone}
There are also numerous reports of teams partying to the detriment of the game.
Within the NBA, J.R. Smith's performance metrics substantially improved 
 after being traded from New York to Cleveland; when an interviewer
asked what contributed to this bump in performance, he infamously replied
that in New York he would have been out partying, but because there's no nightlife in Cleveland he ends up hitting the gym.\citep{price,ley} Before being traded, J.R. Smith's career-to-date success rate of shots taken was 42 percent, but somehow he only managed 
34 percent on Sunday afternoon games in the 2012-13 season.
Presumably, this was because such games followed a long weekend of partying.\citep{princeofthecity}

\subsection{Individual players}
Studies have also been carried out on statistical ranking of the players
most notorious for pursuing active nightlife in the NBA.\citep{chase} In particular, 
the author rejects the null hypothesis that J.R. Smith shoots just as well on Sundays as he does
every other day of the week at the $p=0.01$ level of significance; however, the author also
cautions that with so many players in the NBA, it is expected that we find a split as extreme
as J.R.'s simply due to chance.
The study goes on to look
at historical performance for all NBA players who began their career after 1974, and compares
Sunday afternoon performance to overall career performance. A binomial test is carried out, and the 
study concludes with strong statistical significance that across the NBA, 
individual players perform worse on Sundays when compared with other days, 
as measured by the ratio of shots scored to shots taken.

\subsection{Teams}
There has been an attempt to answer the question at the team-level before: i.e. do teams perform
worse after a night on the town, and if so does it depend on which city they 
visited.\citep{ezekowitz} However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller, because the author only considered Sunday afternoon games
from 2006-14. Finally, there was no methodology used behind the selection of party cities:
the authors first tried to define ad-hoc five ``party cities'', and when that did not yield
a significant result they tried estimating fixed-effects for each city. The author does carry out a binomial test of how many home-game covers vs. away-game covers were met, 
and concludes
with weak statistical evidence that teams systematically play worse on Sunday afternoons 
when on the road, regardless of city.

>>>>

\section{Introduction}
It's not unreasonable to think that in-game sporting performance can be affected 
partly by what takes place off the court. 
We can't observe what happens between games directly. Instead, we proxy for the 
possibility of athletes partying by looking at play following games in party cities.
We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife; we call this a ``hangover effect''.
Part of the question is determining a reasonable way to measure
levels of nightlife, and correspondingly which cities are notorious for it; we colloquially refer to such cities as ``party cities''.
At first glance, the topic might seem callow, but in fact there are many interesting
nuances to work through.

Inquiring about this causal relationship is worthwhile for several reasons. 
Knowing how player performance is affected by external factors could inform coaches of a more
optimal play strategy: if visitation to a city with active nightlife is correlated with
next-day injury, perhaps coaches might bench their star players if the latter game isn't crucial.
It could also inform team managers how travel schedules should be set, or even team rules.
Further, there are potential opportunities for arbitrage.  The
NBA Commissioner Adam Silver has reported that there is \$400 billion per year spent 
on sports betting in the U.S. alone.\citep{asilver,cnnsilver}

% \paragraph{What do we do}
How do we carry out this study?
We exploit data on bookmaker spreads: the expected score 
differential between two teams after conditioning on observable performance in past games
and expectations about the upcoming game. 
We expect a team to meet the spread half the time, since this is one of the easiest ways for 
bookmakers to guarantee a profit.\citep{dubnerlevitt} We construct a model which attempts to 
estimate the causal effect of  visiting a ``party city'' on subsequent day 
performance as measured by the odds of beating the spread. In particular, we only consider the hangover effect 
on games played back-to-back within 24 hours of each other.
To the extent that odds of beating the spread against next day opponent 
is uncorrelated with playing in a party city the day before, which should be the case 
under an efficient betting market, we have identification in our variable of interest.

% \paragraph{What do we find}
We perform this analysis for both the National Basketball Association
(NBA) and Major League Baseball (MLB). 
We find that visiting a city with active nightlife the day prior to a game does have a 
statistically significant negative effect on a team's likelihood of meeting bookmakers' expectations
for both NBA and MLB. Within the NBA, we analyze specific team performance metrics such as number of points allowed and number of rebounds obtained in a game, and find that both are negatively effected by previous day visitation to a city with active nightlife. Having established a causal model, we realize a profitable betting scheme for MLB. We also perform a robustness check and see that if players rest more than
 24 hours after visiting a ``party city'', the hangover effect dissipates.

@@

\subsection{Anecdotal evidence of the effects on sporting performance}
Professional players openly admit they are interested in nightlife: Matt Barnes recently complained that 
``there's no nightlife in Utah'' when asked about playing the Jazz in the second round of NBA playoffs 
in 2017.\citep{cestone}
There are also numerous reports of teams partying to the detriment of the game.
Within the NBA, J.R. Smith's performance metrics substantially improved 
 after being traded from New York to Cleveland; when an interviewer
asked what contributed to this bump in performance, he infamously replied
that in New York he would have been out partying, but because there's no nightlife in Cleveland he ends up hitting the gym.\citep{price,ley} Before being traded, J.R. Smith's career-to-date success rate of shots taken was 42 percent, but somehow he only managed 
34 percent on Sunday afternoon games in the 2012-13 season.
Presumably, this was because such games followed a long weekend of partying.\citep{princeofthecity}

\subsection{Individual players}
Studies have also been carried out on statistical ranking of the players
most notorious for pursuing active nightlife in the NBA.\citep{chase} In particular, 
the author rejects the null hypothesis that J.R. Smith shoots just as well on Sundays as he does
every other day of the week at the $p=0.01$ level of significance; however, the author also
cautions that with so many players in the NBA, it is expected that we find a split as extreme
as J.R.'s simply due to chance.
The study goes on to look
at historical performance for all NBA players who began their career after 1974, and compares
Sunday afternoon performance to overall career performance. 
In particular, the study examines
2,035 players with at least one shot on Sunday games and finds that 1,092 of them shot worse
on Sundays than they did on all other days, as measured by the ratio of shots scored to shots
taken. A binomial test concludes that across the NBA, 
individual players perform worse on Sundays. Effect sizes are not considered, 
i.e. although players perform worse, we don't know by how much.

\subsection{Teams}
There has been an attempt to answer the question at the team-level before: i.e. do teams perform
worse after a night on the town, and if so does it depend on which city they 
visited.\citep{ezekowitz} However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, the sample size was much smaller than ours, because the author only considered Sunday afternoon games from 2006-14. Finally, there was no methodology used behind the selection of party cities: the authors first define ad-hoc five ``party cities'', and when that did not yield a significant result they estimate fixed-effects for each city. The author does carry out a binomial test of how many home-game covers vs. away-game covers were met, and concludes with weak statistical evidence that teams 
systematically play worse on Sunday afternoons when on the road, regardless of city, as measured by
their ability to meet bookmakers' expectations.

Allow us to elaborate on why we think that a delayed hangover effect is a more plausible model.
Suppose for a moment that teams do in fact systematically play worse when playing with cities in active nightlife.
Notice that a simple fixed effect for the team match-up and an interaction with a home team indicator would be sufficient to model this.
Given the size of the betting market, it's not hard to believe that such a simple arbitrage opportunity has already been considered
and factored into bookmakers' expectations. There is also the issue of an unknown travel schedule: if a team plays in a party city,
we can't infer whether they even had the option of being exposed to nightlife without looking at where they were the day prior.
This motivates us to examine a feature describing lagged game location.
